---
title: "Hacking Medical Physics with R"
author: "Michael Wieland"
date: "`r Sys.Date()`"
output: 
  html_document: 
    highlight: pygments
    theme: flatly
    toc: yes
    toc_float: yes
    code_folding: show
    keep_md: yes
---

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
library(ggthemes)
library(kableExtra)
library(RSQLite)
```

## Short Description
This is a R Version for Part 2 of the Article Series "Hacking Medical Physics" by Jonas Andersson and Gavin Poludniowski i (GitRepo: [rvbCMTS/EMP-News](https://github.com/rvbCMTS/EMP-News.git)) in the Newsletter of the European Federation of Organisations for Medical Physics (EFOMP)^[[European Medical Physics News](https://www.efomp.org/index.php?r=fc&id=emp-news)].

### Ressources and Preparation
In order to run this R-Markdown file you need to install RStudio/R ([RStudio Installer](https://www.rstudio.com/products/rstudio/download/)).  

Ressources to get started with R:  

* [RStudio Education - Beginners](https://education.rstudio.com/learn/beginner/)  
* [RStudio Support - Getting Started with R](https://support.rstudio.com/hc/en-us/articles/201141096-Getting-Started-with-R)  
* [RStudio Book Collection](https://www.rstudio.com/products/rstudio/download/)  
* [ggplot2 - Elegent Graphics for Data Analysis](https://ggplot2-book-solutions-3ed.netlify.app/index.html)  

I will make heavy use of the package collection `tidyverse` and the "pipe"-operator (` %>% `). To learn more have a look at: [`tidyverse` - R packages for data science](https://www.tidyverse.org/).  
If you have not installed the packages loaded in the `setup code chunk` (see above) start with installing them via `Tools` -> `Install Packages`.

## Reading in Data
If your data files reside in the working directory you can access them in a relative fashion. My current working directory is the folder where this R-Markdown-File is stored and the data files are stored in a subfolder called "reports".

### Reading an Excel File
My preferred method to read Excel files is to use the `readxl`-package:

```{r reading_excel}
read_xls(path = "reports/StaffDoses_1.xls")
```

### Fixing the column names
Reading the Excel-File with the function `read_xls` from the package `readxl` gives a decent first result. A few things should be changed though in order to work with the data properly. The variable names (column titles) should follow the following convention^[[Social Science Computing Cooperative - Naming Variables](https://sscc.wisc.edu/sscc/pubs/DWE/book/4-2-naming-variables.html)]:  

> * Use only lower case.  
> * Use the underscore, "_" as a replacment for spaces to separate words (called __snake coding__).
> * ...

Assuming that the reports are always delivered in the same format and structure we can fix the column headers once and use them later on to replace the column names for all reports.

```{r fixing_report_header}
report_column_names <- read_xls(path = "reports/StaffDoses_1.xls",
         n_max = 0) %>% # to extract the column names we don't need any data 
  colnames() %>% # exctracting the column names as a vector
  tolower() %>% # convert upper case to lower case
  # next we use the function "gsub":
    # https://statisticsglobe.com/sub-gsub-r-function-example
    # https://rstudio-pubs-static.s3.amazonaws.com/74603_76cd14d5983f47408fdf0b323550b846.html
  gsub(pattern = " ", replacement = "_") %>% # replacing blanks with underscores
  gsub(pattern = "[().]", replacement = "")  # replacing round brackets and dots; 
    # the square-brackets function as list operator (all characters inside the square-brackets are identified)

report_column_names
```

### Read in all reports from a folder
To read in all files from a folder we can make use of the function `read.files()` that gives a list of all files in a folder.

```{r read_all_files_from_folder}
list.files("reports") # get a list of all files from a folder
all_reports_to_read_in <- list.files("reports")

# number of reports in the folder:
length(all_reports_to_read_in)

all_reports <- data.frame() # create a dataframe to hold all reports

for (i in 1:length(all_reports_to_read_in)) { # a for-loop to read in all reports
  rep <- read_xls(path = paste0("reports/", all_reports_to_read_in[i])) # reading in the i-th report into veriable "rep"
  all_reports <- rbind(all_reports, rep) # binding together the reports rowwise
}  

colnames(all_reports) <- report_column_names # replacing the column names with the fixed names (see above)
```

### Fix data types
Some data wrangling is needed to get the right data types: 

* All numerical variables should be defined as `double` or `integer`,  
* Replace "," with "." in decimal numbers so R can recognise them as numbers ("English convention" for decimal numbers),  
* Create new columns `hp10_status` and `hp007_status` before converting `hp10` and `hp007` to numeric in order not to lose information. Where `hp10` and `hp007` have the values B, NR or `NA` (B: Below Measurement Treshold; NR: Not returned; `NA`: Missing Value) we transfer those values to the new columns, if the values are numeric we set the value in the new columns to OK.  

To fix the dates I needed a work around because my machine `locale` is set to German but the dates in the reports have abbreviated month names in English. One way to read in the data correctly with little coding is to set the `locale` on the machine to English temporarily.



```{r fix_data_types, warning=FALSE}
# getting locale
loc <- Sys.getlocale("LC_TIME")
Sys.setlocale("LC_TIME", locale = "English") 
# the value for "locale" is depending on the operating system. For Windows the value is "English".
# check the help page with "?Sys.setlocale()" if you are not using Windows.


all_reports_fixed <- all_reports %>% 
  # In first step we are replacing the colon with dots as comma sign (needed to convert to numeric later on)
  # with the function mutate we create new columns based on existing ones or modify the content of existing columns
  mutate(hp10 = str_replace_all(hp10, pattern = ",", replacement = ".")) %>% 
  mutate(hp007 = str_replace_all(hp007, pattern = ",", replacement = ".")) %>% 
  # before we convert hp10 and hp007 to numeric we create new columns with non-numeric values of hp10 and hp007 in order not to lose information
  # "B": Below Measurement Treshold, "NR": Not returned, "NA": missing value; "OK" where a numeric value exisists for the variables hp10 and hp007)
  mutate(hp10_status = case_when(hp10 == "B" ~ "B", 
                                 hp10 == "NR" ~ "NR",
                                 is.na(hp10) ~ NA_character_, 
                                 is.numeric(as.numeric(hp10)) ~ "OK"),
         hp007_status = case_when(hp007 == "B" ~ "B", 
                                 hp007 == "NR" ~ "NR",
                                 is.na(hp007) ~ NA_character_, 
                                 is.numeric(as.numeric(hp007)) ~ "OK")) %>% 
  # next we convert relevant columns to numeric (non-numeric values in hp10 and hp007 will be converted to NA automatically)
  mutate(across(c(customer_uid, department_uid, person_uid, hp10, hp007, dosimeter_uid, report_uid), as.numeric)) %>% 
  # to make sure we have no duplicated data (same report read in more than once, identified by "report_uid") 
  # we eliminate duplicates with the function "distinct"... after grouping by person_uid and dosimeter_uid
  mutate(across(c(measurement_period_start:report_date), as.Date, format = "%d-%b-%Y")) %>% 
  group_by(person_uid, dosimeter_uid) %>% 
  distinct(report_uid, .keep_all = TRUE) %>%
  ungroup()

head(all_reports_fixed)

Sys.setlocale("LC_TIME", locale = loc) # setting back the locale 
```

#--------  
__Comment Michael: Missunderstanding regarding Status column -> Change to be equivalent to Python tutorial__  
#--------


## Reporting

### Hp(10) - Summary Statistics by Department and Year
```{r hp10_summary_stat}
all_reports_fixed %>% 
  drop_na(hp10) %>%  # droping all rows where no dose was recorded (not really necessary, ggplot can handle NAs)
  mutate(report_year = format(report_date, "%Y")) %>% # extracting the year from the report date
  filter(report_year %in% c("2020", "2021")) %>% # chose years 2020 and 2021
  ggplot(aes(x=report_year, y=hp10)) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0,NA), # set lower bound to 0 and let ggplot automatically set upper bound
                     breaks = pretty(c(0, max(all_reports_fixed$hp10, na.rm = T)), n=10)) + # getting 10 breaks in the y-axis
  labs(x = "", y = "Hp(10) [mSv]", # Axis names
       title = "Summary Statistics for monthly Hp(10) values",
       subtitle = "  by department and year",
       caption = "Hacking Medical Physics. Data and Idea: J. Andersson and G. Poludniowski") +
  # tweaking the plot style (optional) with the function "theme":
  theme(panel.background = element_blank(),
        panel.grid.major.y = element_line(linetype = 3, color = "gray"),
        panel.spacing.x = unit(1, "cm"),
        strip.text = element_text(face = "bold", color = "white"),
        strip.background = element_rect(fill = "#D47B60")) +
  facet_wrap(~ department) # split up the data into subplots for the different departments
```

### Number of Staff Dose Readings per Dosimeter Type
```{r nr_dose_readings}
all_reports_fixed %>% 
  filter(user_type == "Staff") %>% # Filter for Staff readings
  filter(! hp007_status %in% c("NR", "B")) %>% 
  # grouping by dosimeter type to count both types seperately
  group_by(dosimeter_type) %>% # to count dosimeter readings depending on type
  summarise(sum(!is.na(hp10)), length(!is.na(hp007))) %>% # counting the rows with numeric values only (that are not NA)
    # OBS: !is.na(a) gives back TRUE if a variable is NOT NA. In R the logical value TRUE is aquivalent to 1. 
    # Therefore summing over a vector of logical values gives the number of "TRUE".
  kable(align = "lcc", # kable is a function to produce tables
        col.names = c("Dosimeter Type", "Hp(10)", "Hp(0.07)"), 
        caption = "Number of dosimeter readings per dosimeter type") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) # a function to further tweak tablbes
```


### Individuals who did not return their Dosimeters
```{r not returned}
all_reports_fixed %>% 
  filter(hp007_status == "NR") %>% 
  count(name) %>% 
  kable(align = "lc", # kable is a function to produce tables
        col.names = c("Name", "Instances"), 
        caption = "Number of times a dosimeter was not returned per person") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) # a function to further tweak tablbes
```


## Using R with SQL

### Ressources and Motivation
For this part I am drawing heavily on the following two ressources:  

* [RStudio - Databases using R](https://db.rstudio.com/)  
* [datacamp.com - SQLite in R](https://www.datacamp.com/community/tutorials/sqlite-in-r)  

For a limited number of files like in the example above working with a database is not necessary but databases have several advantages^[[opentextbc.ca - Database Design](https://opentextbc.ca/dbdesign01/chapter/chapter-3-characteristics-and-benefits-of-a-database)]:  

>* Data Indepence (your colleagues might want to access the data with Python or Matlab)  
>* Insulation between data and program  
>* Support for multiple views of data (subsets of the data for different users)  
>* Centralized control over data  
>* Data can be shared  
>* Redundancy can be reduced (ideally each data item is stored in only one place)  
>* Integrity constraints (rules that dictate what can be entered or edited)  
>* Security constraints  

### Getting started
If your data is not already available in a database you have to create one:

```{r creating_a_db}

```


## Reporting with RMarkdown


#-------------
```{r test, eval=FALSE, echo=FALSE}
# Testing SQLite with datacamp tutorial
# https://www.datacamp.com/community/tutorials/sqlite-in-r

data("mtcars")
mtcars$car_names <- rownames(mtcars)
rownames(mtcars) <- c()
head(mtcars)
# Create a connection to our new database, CarsDB.db
# you can check that the .db file has been created on your working directory
connection <- dbConnect(RSQLite::SQLite(), "CarsDB.db")

# Write the mtcars dataset into a table names mtcars_data
dbWriteTable(conn = connection, name = "cars_data", value = mtcars)
# List all the tables available in the database
dbListTables(conn)


# appending data from different files
# Create toy data frames
car <- c('Camaro', 'California', 'Mustang', 'Explorer')
make <- c('Chevrolet','Ferrari','Ford','Ford')
df1 <- data.frame(car,make)
car <- c('Corolla', 'Lancer', 'Sportage', 'XE')
make <- c('Toyota','Mitsubishi','Kia','Jaguar')
df2 <- data.frame(car,make)
# Add them to a list
dfList <- list(df1,df2)
# Write a table by appending the data frames inside the list
for(k in 1:length(dfList)){
    dbWriteTable(connection, "Cars_and_Makes", dfList[[k]], append = TRUE)
}
# List all the Tables
dbListTables(connection)

# check to see if data is in table
dbGetQuery(connection, statement = "SELECT * FROM Cars_and_Makes")
```

```{r test2, eval=FALSE, echo=FALSE}
# Query examples
dbGetQuery(connection, "SELECT * from cars_data LIMIT 5")
dbGetQuery(connection, "SELECT car_names, mpg, cyl, disp from cars_data WHERE gear = 4")
dbGetQuery(connection, "SELECT car_names, hp, cyl FROM cars_data WHERE car_names LIKE 'M%' AND cyl IN (6,8)")
dbGetQuery(connection, "SELECT cyl, AVG(hp) AS 'average_hp', AVG(mpg) AS 'average_mpg' FROM cars_data
           GROUP BY cyl
           ORDER BY average_hp")
```

```{r test3, eval=FALSE, echo=FALSE}
# Close the database connection to CarsDB
dbDisconnect(conn)
```

```{r test4, eval=FALSE, echo=FALSE}

```

```{r test5, eval=FALSE, echo=FALSE}

```